Comparing Different Machine Learning Models at Predicting Bitcoin Price Moving Direction at Different Time Intervals from 2017 to 2024

### 2024-12-05 paper/old paper.qmd
EDA graphs saved. Abstract, Intro, Data completed. LFS tracks all parquet files now.

---
title: "Comparing Different Machine Learning Models at Predicting Bitcoin Price Moving Direction at Different Time Intervals from 2017 to 2024"
subtitle: "RNN Can Predict Best at One-Hour Intervals"
author: 
  - Justin (Jiazhou) Bi
thanks: "All project related files available at: https://github.com/Jiazhou-Bi/BTC-Price-Prediction"
date: Nov 20, 2024
date-format: long
abstract: "This project aims to build a predictive model for the price moving direction of Bitcoin (BTC) at 1-minute, 1-hour, and 1-day intervals. We explored four popular machine learning algorithms for time-series data prediction: Recurrent Neural Network (RNN), Long Short-Term Memory (LSTM), Random Forest (RF), and Extreme Gradient Boosting (xGB). We have compared these four models on two different dimensions: model performance and time for training. The data used for this project is gathered from Binance which contains BTC data from late 2017 to present (November 2024).The original variables include high, low, open, close, and volume, as well as the timestamp of BTC, at one-minute, one-hour, and one-day intervals. Ultimately, RNN has the overall best performance in terms of its predictive accuracy, while XGB has the fastest training duration. While most models across different time intervals achieved a better performance than simply guessing at a 50-50 percent chance, their performance varies for each direction of the movemenet. Future steps for improving the model may involve incorporating additional features to enhance model accuracy and reduce errors. Additionally, fine-tuning hyperparameters, experimenting with more sophisticated feature engineering, and handling outliers could further improve performance."

@@

Cryptocurrencies have drawn much attention of the public recently due to its growing public awareness. xxx amount of papers/news about BTC was published (refereces here) from xxxx to xxxx. According to xxx (reference here), the value of the market... etc. The trading volume (references) ... It has been adopted by xxx platforms and .... (reference) Also, mention the importance of BTC (references needed) and its role of being a indicator of the overall crypto market. Therefore, it has become one of the most traded ....
On the other hand, due to its non-tangible nature, volitility is key aspect. xxxx data about its ups and downs. Only contains data from 2017 because .... fits our scope. Regain values (references 5 in the paper).
Similar to the stock prices, because BTC is considered a tradable investment instrument, many attempt to predict its price in the future to faciliate their trading strategy. Recently studies were published to discuss whether the price of BTC is predictable and technical indicators returned the possibility. Accuracy up to 63% for direction prediction (reference here, 9). Also, in 10, they reported 98% success rate for daily price forecast. Based on recent data 2017 and 2018. Both EOD prediction. We have even short trading horizons, 1 min, 1 hour.

Income is a fundamental concept as it affects nearly every member of society, influencing access to resources, quality of life, and overall economic stability. At the same time, it is also a difficult concept to define, as it can be interpreted differently depending on various perspectives, such as economic, social, legal, and tax purposes. Therefore, we must examine various demographical factors that may impact income to understand its concept better. Our project defines income as "all forms of net financial resources generated or lost from work, investments, or other activities." This project aims to predict personal income using data from the 2023 American Community Survey (ACS) [@ipums_usa_v15]. The ACS provides detailed demographic information, including age, education, employment, marital status, mortgage status, veteran status. As a result, it provides a robust dataset for studying potential factors that may influence income.

Due to the timeframe limitations, the findings of this project are specifically applicable to the 2023 dataset. As such, the results may vary across different periods, and their generalizability has not been assessed within this project. The raw data used in this analysis was sourced from IPUMS using the IPUMS API [@ipums_api]. We utilized pandas [@pandas] for data processing, and data visualization was carried out using Seaborn [@seaborn]. Additionally, all tables presented in this report were generated with Plotly [@plotly].

Income inequality is a serious problem and can lead to significant consequences and it may affect social structure, economic stability, as well as policy-making processes. Researchers found that demographic factors, such as education level and occupation type, can significantly impact an individual's financial standing [@Autor_2013]. Similarly, social relationships, such as marital status and household characteristics, affect income dynamics [@WaiteGallagher_2000]. Therefore, applying machine learning models to analyze the ACS dataset and using the information mentioned to predict personal income is reasonable.

>>>>

---
title: "Comparing Different Machine Learning Models at Predicting Bitcoin Price Moving Direction at Different Time Intervals from 2017 to 2024"
subtitle: "RNN Can Predict Best at One-Hour Intervals"
author: 
  - Justin (Jiazhou) Bi
thanks: "All project related files available at: https://github.com/Jiazhou-Bi/BTC-Price-Prediction"
date: Dec 3, 2024
date-format: long
abstract: "This project aims to build a predictive model for the price moving direction of Bitcoin (BTC) at 1-minute, 1-hour, and 1-day intervals. We explored four popular machine learning algorithms for time-series data prediction: Recurrent Neural Network (RNN), Long Short-Term Memory (LSTM), Random Forest (RF), and Extreme Gradient Boosting (xGB). We have compared these four models on two different dimensions: model performance and time for training. The data used for this project is gathered from Binance which contains BTC data from late 2017 to present (November 2024).The original variables include high, low, open, close, and volume, as well as the timestamp of BTC, at one-minute, one-hour, and one-day intervals. Ultimately, RNN has the overall best performance in terms of its predictive accuracy, while XGB has the fastest training duration. While most models across different time intervals achieved a better performance than simply guessing at a 50-50 percent chance, their performance varies for each direction of the movemenet. Future steps for improving the model may involve incorporating additional features to enhance model accuracy and reduce errors. Additionally, fine-tuning hyperparameters, experimenting with more sophisticated feature engineering, and handling outliers could further improve performance."

@@

Cryptocurrencies have drawn much attention of the public recently due to its growing public awareness. Researchers have found that 4,495 literatures related to Bitcoin were published from 2011 to 2020 [@aysan2021ascent]. More recently, another study stated that more than 9,100 research papers were published on cryptocurrencies in the Web of Science [@bara2024impact]. In 2015, a journal called Ledger was published by the UNiversity of Pittsburgh's University Library System, and it is the first journal dedicated to cryptocurrencies [@extance2015bitcoin].This reflects the trend of growing interest in digital asset. In December, 2019, the were 2,813 different cryptocurrencies exsiting with a 201.47 billion US dollars in their total market capitalization; in particular, bitcoin is the biggest contributor and had a 131.89 billion US dollars market capitalizaiton [@chen2021empirical]. As I am writing this article on December 4, 2024, the current market capitalization of Bitcoin is 1.95 trillion US dollars, with a 103.22 billion US dollars 24-hour trading volume, accroding to forbes.com [@Forbes]. As of October 2023, there were a total of 1492 cryptocurrency trading plaforms globally, both centralized and decentralized [@Oh_2024]. Among all the different cryptocurrencies, Bitcoin is the key player in both the market and in the research field [@aysan2021ascent]. Therefore, learning more about bitcion can be benificial for both the reserachers, and the public who has interest in cryptocurrency and their tradings.

On the other hand, due to its non-tangible nature, volatility is a key aspect of Bitcoin. Much reserach has focused on the price fluctuations of Bitcoin, and compared its volatility to more traditional assets, such as gold. For example, [@kurihara2018does] demostrated that Bitcoin's higher volatility compared to gold, indicating its speculative nature. In another study, the authors have concluded that Bitcoin's extreme price volatility may interfere with its use as a stable investment instrument. As a result, speculative investors are often more attracted to Bitcoin [@chen2021empirical], and looking for profit by capitalizing on its rapid price changes [@baur2018bitcoin].

Despite its extreme volatility, Bitcoin is still widely accepted as a tradable investment vehicle, as many of characteristics are similar to stocks [@chen2021empirical]. In a very recently published study, the authors developed a model called TradExpert-NM, and it can achieve a 59% accuracy for Bitcoin's daily price movement direction [@ding2024tradexpert]. In another study, [@mudassir2020time] developed a model that can score up to 65% in accuracy for next-day forecast of Bitcoin's price movement. However, many of these "high-performing" models are using various data (social media data in [@ding2024tradexpert] and other technical data [@mudassir2020time]) that are usually too complicated for investors with limited knowledge in machine learning and/or data science. Even for models used only more accessible data, such as in [@adcock2019non], they utilized neural network as their model.
